---
title: Memory decline in elderly with cerebral small vessel disease explained by temporal
  interactions between white matter hyperintensities and hippocampal atrophy
output:
  html_document:
    df_print: paged
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F,
                      message = F,
                      warning = F,
                      results='asis')

library(stargazer)
library(lme4)
library(tidyverse)

df = read.csv("../data/RUNDMC_data_long.csv", sep=";", dec=",")

# apply data transformations

df$age06s <- scale(df$age06)
df$sex <- factor(df$sex)
df$rundmcs <- factor(df$rundmcs)
df$timesq <- df$time^2
df$lnwmh <- log(df$wmh)
altIndex <- c("psychomotorspeed", "fluency", "responseinhibition", "attention",
              "executivefunction")
df$cognitiveindex_nomem <- rowMeans(df[altIndex], na.rm=T)
df$gmvnohv <- (df$gmv - df$hv)
df$em <- (df$immediatememory + df$delayedmemory)/2
df <- df[complete.cases(df[, c("lnwmh", "hv")]),]

```

## 1. Showing the relationship between age and memory decline

First, we created a "null" model, which expressed the effect of age and time on a composite memory score. Years of education and sex were also added to the model as static covariates. In this first model, baseline age represents the cross-sectional effects of age on memory, while time between follow-ups and the square of time between follow-ups represent the linear and quadratic effects of temporal progression on memory, respectively. 


```{r}
model1 <- lme4::lmer(memory ~ 
                       age06 + 
                       time + 
                       timesq +
                       sex + 
                       educationyears +
                     (1 + time|rundmcs),
                     data=df, REML=F)
stargazer(model1, header=F, type="html",  
          ci=T, ci.level = .95,
          omit="Constant",
          order=c(1,4,5,2,3),
          covariate.labels = c("Baseline age (years)", 
                               "Sex",
                               "Education (years)",
                               "Time to follow-up (linear)", 
                               "Time to follow-up (quadratic)"))

```


## 2. Examining the interaction between WMH and HV in explaining memory decline

We then adopted a data-driven approach to determine the role of WMH and HV in contributing to memory deficits. Using the null model as a baseline, three successive models were created: one with only the effects of WMH, a second with only the effects of HV, and a third with the simulatenous effects of WMH and HV. 

These three models, along with the null model, were then compared to see which model best explained the data. Since the random effect is identical across all models, the fixed effects are compared. Also all other models can be seen as restricted cases of the final model. 

In order to facilitate comparisons between models with different fixed effects, models were fit by minimising the negative log-likelihood. All models were then compared using a one-way ANOVA.


```{r}

model2 <- lme4::lmer(memory ~ age06 + time + timesq*lnwmh +
                     sex + educationyears +
                     (1 + time|rundmcs), 
                     data=df, REML=F)

model3 <- lme4::lmer(memory ~ age06 + sex + time*hv + timesq  +
                     sex + educationyears +
                     (1 + time|rundmcs),
                     data=df, REML=F)

model4 <- lme4::lmer(memory ~ age06 + time*hv + timesq*lnwmh +
                     hv*lnwmh  + sex + educationyears +
                     (1 + time|rundmcs),
                     data=df, REML=F)

model5 <- lme4::lmer(memory ~ age06 + time*hv + timesq*lnwmh +
                     sex + educationyears +
                     (1 + time|rundmcs),
                     data=df, REML=F)

stargazer(model1, model2, model3, model4, 
          title = "Fixed effects results",
          ci=T, ci.level = .95, 
          omit="Constant", dep.var.labels = "Memory",
          order=c(1,6,7,2,4,5,8,3,10,9),
          covariate.labels = c("Baseline age (years)", 
                               "Sex",
                               "Education (years)",
                               "Time to follow-up (linear)", 
                               "Time to follow-up (quadratic)", 
                               "WMH", 
                               "WMH progression",
                               "HV", 
                               "Hippocampal atrophy", 
                               "WMH * HV interaction"),
          star.cutoffs = c(0.05, 0.01, 0.001),
          column.labels = c("Null", "WMH only", "HV only", "Full"),
          initial.zero = F,
          header=F, type="html")

stargazer(anova(model1, model2, model3, model4), summary = F, type="html")

stargazer(anova(model5, model4), summary = F, type="html")

```

Both models with WMH only and HV only provided significantly better fit in comparison to the null model. The model with HV alone fit better than the model with WMH alone. Importantly, however, the model with both WMH and HV provided the best fit to the data. 

## 3. Memory examined separately as episodic memory (i.e. immediate and delayed memory) and working memory

Working memory and episodic memory were examined separately, as we have indications that they are affected differently by WMH and HV [^1].

### 3a. Working memory

Here we examined Working Memory separately.

```{r}
wm.model1 <- lme4::lmer(workingmemory ~ 
                       age06 + 
                       time + 
                       timesq +
                       sex + 
                       educationyears +
                     (1 + time|rundmcs),
                     data=df, REML=F)

wm.model2 <- lme4::lmer(workingmemory ~ age06 + time + timesq*lnwmh +
                     sex + educationyears +
                     (1 + time|rundmcs), 
                     data=df, REML=F)

wm.model3 <- lme4::lmer(workingmemory ~ age06 + sex + time*hv + timesq  +
                     sex + educationyears +
                     (1 + time|rundmcs),
                     data=df, REML=F)

wm.model4 <- lme4::lmer(workingmemory ~ age06 + time*hv + timesq*lnwmh +
                     hv*lnwmh  + sex + educationyears +
                     (1 + time|rundmcs),
                     data=df, REML=F)

wm.model5 <- lme4::lmer(workingmemory ~ age06 + time*hv + timesq*lnwmh +
                     sex + educationyears +
                     (1 + time|rundmcs),
                     data=df, REML=F)

stargazer(wm.model1, wm.model2, wm.model3, wm.model4, 
          title = "Fixed effects results",
          ci=T, ci.level = .95, 
          omit="Constant", dep.var.labels = "Working Memory",
          order=c(1,6,7,2,4,5,8,3,10,9),
          covariate.labels = c("Baseline age (years)", 
                               "Sex",
                               "Education (years)",
                               "Time to follow-up (linear)", 
                               "Time to follow-up (quadratic)", 
                               "WMH", 
                               "WMH progression",
                               "HV", 
                               "Hippocampal atrophy", 
                               "WMH * HV interaction"),
          star.cutoffs = c(0.05, 0.01, 0.001),
          column.labels = c("Null", "WMH only", "HV only", "Full"),
          initial.zero = F,
          header=F, type="html")

stargazer(anova(wm.model5, wm.model4), summary = F, type="html")

```

### 3b. Episodic memory

Here we examined Episodic Memory separately.

```{r}
em.model1 <- lme4::lmer(em ~ 
                       age06 + 
                       time + 
                       timesq +
                       sex + 
                       educationyears +
                     (1 + time|rundmcs),
                     data=df, REML=F)

em.model2 <- lme4::lmer(em ~ age06 + time + timesq*lnwmh +
                     sex + educationyears +
                     (1 + time|rundmcs), 
                     data=df, REML=F)

em.model3 <- lme4::lmer(em ~ age06 + sex + time*hv + timesq  +
                     sex + educationyears +
                     (1 + time|rundmcs),
                     data=df, REML=F)

em.model4 <- lme4::lmer(em ~ age06 + time*hv + timesq*lnwmh +
                     hv*lnwmh  + sex + educationyears +
                     (1 + time|rundmcs),
                     data=df, REML=F)

em.model5 <- lme4::lmer(em ~ age06 + time*hv + timesq*lnwmh +
                     sex + educationyears +
                     (1 + time|rundmcs),
                     data=df, REML=F)

stargazer(em.model1, em.model2, em.model3, em.model4, 
          title = "Fixed effects results",
          ci=T, ci.level = .95, 
          omit="Constant", dep.var.labels = "Episodic Memory",
          order=c(1,6,7,2,4,5,8,3,10,9),
          covariate.labels = c("Baseline age (years)", 
                               "Sex",
                               "Education (years)",
                               "Time to follow-up (linear)", 
                               "Time to follow-up (quadratic)", 
                               "WMH", 
                               "WMH progression",
                               "HV", 
                               "Hippocampal atrophy", 
                               "WMH * HV interaction"),
          star.cutoffs = c(0.05, 0.01, 0.001),
          column.labels = c("Null", "WMH only", "HV only", "Full"),
          initial.zero = F,
          header=F, type="html")

stargazer(anova(em.model5, em.model4), summary = F, type="html")

```

### Working memory vs. Episodic memory

```{r}

MemLabel <- c("Working Memory", "Episodic Memory")

stargazer(wm.model1, wm.model2, wm.model3, wm.model4, em.model1, em.model2, em.model3, em.model4, 
          title = "Fixed effects results",
          ci=T, ci.level = .95, 
          omit="Constant", dep.var.labels = MemLabel,
          order=c(1,6,7,2,4,5,8,3,10,9),
          covariate.labels = c("Baseline age (years)", 
                               "Sex",
                               "Education (years)",
                               "Time to follow-up (linear)", 
                               "Time to follow-up (quadratic)", 
                               "WMH", 
                               "WMH progression",
                               "HV", 
                               "Hippocampal atrophy", 
                               "WMH * HV interaction"),
          star.cutoffs = c(0.05, 0.01, 0.001),
          column.labels = c("Null", "WMH only", "HV only", "Full", "Null", "WMH only", "HV only", "Full"),
          initial.zero = F,
          header=F, type="html")

stargazer(anova(wm.model5, wm.model4), summary = F, type="html")
stargazer(anova(em.model5, em.model4), summary = F, type="html")

```


### 3c. Immediate memory

Here we examined Immediate Memory separately.

```{r}
im.model1 <- lme4::lmer(immediatememory ~ 
                       age06 + 
                       time + 
                       timesq +
                       sex + 
                       educationyears +
                     (1 + time|rundmcs),
                     data=df, REML=F)

im.model2 <- lme4::lmer(immediatememory ~ age06 + time + timesq*lnwmh +
                     sex + educationyears +
                     (1 + time|rundmcs), 
                     data=df, REML=F)

im.model3 <- lme4::lmer(immediatememory ~ age06 + sex + time*hv + timesq  +
                     sex + educationyears +
                     (1 + time|rundmcs),
                     data=df, REML=F)

im.model4 <- lme4::lmer(immediatememory ~ age06 + time*hv + timesq*lnwmh +
                     hv*lnwmh  + sex + educationyears +
                     (1 + time|rundmcs),
                     data=df, REML=F)

im.model5 <- lme4::lmer(immediatememory ~ age06 + time*hv + timesq*lnwmh +
                     sex + educationyears +
                     (1 + time|rundmcs),
                     data=df, REML=F)

stargazer(im.model1, im.model2, im.model3, im.model4, 
          title = "Fixed effects results",
          ci=T, ci.level = .95, 
          omit="Constant", dep.var.labels = "Immediate Memory",
          order=c(1,6,7,2,4,5,8,3,10,9),
          covariate.labels = c("Baseline age (years)", 
                               "Sex",
                               "Education (years)",
                               "Time to follow-up (linear)", 
                               "Time to follow-up (quadratic)", 
                               "WMH", 
                               "WMH progression",
                               "HV", 
                               "Hippocampal atrophy", 
                               "WMH * HV interaction"),
          star.cutoffs = c(0.05, 0.01, 0.001),
          column.labels = c("Null", "WMH only", "HV only", "Full"),
          initial.zero = F,
          header=F, type="html")

stargazer(anova(im.model5, im.model4), summary = F, type="html")

```


### 3d. Delayed memory

Here we examined Delayed Memory separately.

```{r}
dm.model1 <- lme4::lmer(delayedmemory ~ 
                       age06 + 
                       time + 
                       timesq +
                       sex + 
                       educationyears +
                     (1 + time|rundmcs),
                     data=df, REML=F)

dm.model2 <- lme4::lmer(delayedmemory ~ age06 + time + timesq*lnwmh +
                     sex + educationyears +
                     (1 + time|rundmcs), 
                     data=df, REML=F)

dm.model3 <- lme4::lmer(delayedmemory ~ age06 + sex + time*hv + timesq  +
                     sex + educationyears +
                     (1 + time|rundmcs),
                     data=df, REML=F)

dm.model4 <- lme4::lmer(delayedmemory ~ age06 + time*hv + timesq*lnwmh +
                     hv*lnwmh  + sex + educationyears +
                     (1 + time|rundmcs),
                     data=df, REML=F)

dm.model5 <- lme4::lmer(delayedmemory ~ age06 + time*hv + timesq*lnwmh +
                     sex + educationyears +
                     (1 + time|rundmcs),
                     data=df, REML=F)

stargazer(dm.model1, dm.model2, dm.model3, dm.model4, 
          title = "Fixed effects results",
          ci=T, ci.level = .95, 
          omit="Constant", dep.var.labels = "Delayed Memory",
          order=c(1,6,7,2,4,5,8,3,10,9),
          covariate.labels = c("Baseline age (years)", 
                               "Sex",
                               "Education (years)",
                               "Time to follow-up (linear)", 
                               "Time to follow-up (quadratic)", 
                               "WMH", 
                               "WMH progression",
                               "HV", 
                               "Hippocampal atrophy", 
                               "WMH * HV interaction"),
          star.cutoffs = c(0.05, 0.01, 0.001),
          column.labels = c("Null", "WMH only", "HV only", "Full"),
          initial.zero = F,
          header=F, type="html")

stargazer(anova(dm.model5, dm.model4), summary = F, type="html")

```

In all 3 memory domains, the full model with the WMH-HV interaction term showed significantly better model fit as well as a significant WMH-HV interaction term. 
The only difference was that in the model of Working Memory, the main effects of HV and hippocampal atrophy were not significant anymore.


## 4. Determine the specificity of the effect - Global brain atrophy

We then compared this model with several alternative competing models to determine the specificity of the effect.


### 4.1 Grey matter atrophy

We wanted to show that the effect of hippocampal atrophy was not part of general grey matter atrophy.


```{r}

gm.model3 <- lme4::lmer(memory ~ age06 + sex + time*gmvnohv + timesq  +
                     sex + educationyears +
                     (1 + time|rundmcs),
                     data=df, REML=F)

gm.model4 <- lme4::lmer(memory ~ age06 + time*gmvnohv + timesq*lnwmh +
                     gmvnohv*lnwmh  + sex + educationyears +
                     (1 + time|rundmcs),
                     data=df, REML=F)

gm.model5 <- lme4::lmer(memory ~ age06 + time*gmvnohv + timesq*lnwmh +
                     sex + educationyears +
                     (1 + time|rundmcs),
                     data=df, REML=F)

gm.m5 <- lme4::lmer(memory ~ age06 + time*gmvnohv + timesq*lnwmh +
                     sex + (1 + time|rundmcs),
                     data=df, REML=F)

gm.m4 <- lme4::lmer(memory ~ age06 + time*gmvnohv + timesq*lnwmh +
                     gmvnohv*lnwmh  + sex + (1 + time|rundmcs),
                     data=df, REML=F)

stargazer(model1, model2, gm.model3, gm.model4, 
          title = "Fixed effects results",
          ci=T, ci.level = .95, 
          omit="Constant", dep.var.labels = "Memory",
          order=c(1,6,7,2,4,5,8,3,10,9),
          covariate.labels = c("Baseline age (years)", 
                               "Sex",
                               "Education (years)",
                               "Time to follow-up (linear)", 
                               "Time to follow-up (quadratic)", 
                               "WMH", 
                               "WMH progression",
                               "GMV", 
                               "GM atrophy", 
                               "WMH * GMV interaction"),
          star.cutoffs = c(0.05, 0.01, 0.001),
          column.labels = c("Null", "WMH only", "GMV only", "Full"),
          initial.zero = F,
          header=F, type="html")

stargazer(anova(model4, gm.model4), summary = F, type="html")
stargazer(anova(gm.model5, gm.model4), summary = F, type="html")
stargazer(anova(gm.m5, gm.m4), summary = F, type="html")

```

The effect was not specific to hippocampal atrophy, as the interaction term of WMH with GMV was also significantly associated with memory performance.


### 4.2 Combined model with both HV & GMV interactions

We next built a combined model with both WMH-HV and WMH-GMV interactions included, to examine the relative strengths of the WMH-HV and WMH-GMV interactions.


```{r}

spec.m.wmh <- lme4::lmer(memory ~ age06 + sex + educationyears +
                          time + timesq + lnwmh + lnwmh:timesq + 
                     (1 + time|rundmcs),
                     data=df, REML=F)

spec.m.hv <- lme4::lmer(memory ~ age06 + sex + educationyears +
                          time + timesq + lnwmh + hv +
                          lnwmh:timesq + hv:time + hv:lnwmh  + 
                     (1 + time|rundmcs),
                     data=df, REML=F)

spec.m.gmv <- lme4::lmer(memory ~ age06 + sex + educationyears +
                           time + timesq + lnwmh + gmvnohv +
                           lnwmh:timesq + gmvnohv:time + gmvnohv:lnwmh +
                     (1 + time|rundmcs),
                     data=df, REML=F)

spec.m.both <- lme4::lmer(memory ~ age06 + sex + educationyears +
                          time + timesq + lnwmh + hv + gmvnohv +
                           lnwmh:timesq + hv:time + gmvnohv:time +
                            hv:lnwmh  + gmvnohv:lnwmh +
                     (1 + time|rundmcs),
                     data=df, REML=F)

stargazer(spec.m.wmh, spec.m.hv, spec.m.gmv, spec.m.both, 
          title = "Fixed effects results",
          ci=T, ci.level = .95, 
          omit="Constant", dep.var.labels = "Memory",
          order=c(1,2,3,4,5,6,9,7,10,11,8,12,13),
          covariate.labels = c("Baseline age (years)", 
                               "Sex",
                               "Education (years)",
                               "Time to follow-up (linear)", 
                               "Time to follow-up (quadratic)", 
                               "WMH", 
                               "WMH progression",
                               "HV", 
                               "Hippocampal atrophy", 
                               "WMH * HV interaction",
                               "GMV",
                               "GM atrophy",
                               "WMH * GM interaction"),
          star.cutoffs = c(0.05, 0.01, 0.001),
          column.labels = c("WMH only", "WMH * HV", "WMH * GMV", "Both"),
          initial.zero = F,
          header=F, type="html")

stargazer(anova(spec.m.wmh, spec.m.hv, spec.m.gmv, spec.m.both), summary = F, type="html")

```

The effect was not specific to hippocampal atrophy, as the interaction term of WMH with GMV was also significantly associated with memory performance. However, when GMV and HV were both included in the model, only the WMH-HV interaction remained significant. 


### 4.3 Combined model with both HV & GMV interactions, seperately for WM vs EM


```{r}

spec.m.wmh.wm <- lme4::lmer(workingmemory ~ age06 + sex + educationyears +
                          time + timesq + lnwmh + lnwmh:timesq + 
                     (1 + time|rundmcs),
                     data=df, REML=F)

spec.m.hv.wm <- lme4::lmer(workingmemory ~ age06 + sex + educationyears +
                          time + timesq + lnwmh + hv +
                          lnwmh:timesq + hv:time + hv:lnwmh  + 
                     (1 + time|rundmcs),
                     data=df, REML=F)

spec.m.gmv.wm <- lme4::lmer(workingmemory ~ age06 + sex + educationyears +
                           time + timesq + lnwmh + gmvnohv +
                           lnwmh:timesq + gmvnohv:time + gmvnohv:lnwmh +
                     (1 + time|rundmcs),
                     data=df, REML=F)

spec.m.both.wm <- lme4::lmer(workingmemory ~ age06 + sex + educationyears +
                          time + timesq + lnwmh + hv + gmvnohv +
                           lnwmh:timesq + hv:time + gmvnohv:time +
                            hv:lnwmh  + gmvnohv:lnwmh +
                     (1 + time|rundmcs),
                     data=df, REML=F)

spec.m.both.wm.contr <- lme4::lmer(workingmemory ~ age06 + sex + educationyears +
                          time + timesq + lnwmh + hv + gmvnohv +
                           lnwmh:timesq + hv:time + gmvnohv:time +
                     (1 + time|rundmcs),
                     data=df, REML=F)


spec.m.wmh.em <- lme4::lmer(em ~ age06 + sex + educationyears +
                          time + timesq + lnwmh + lnwmh:timesq + 
                     (1 + time|rundmcs),
                     data=df, REML=F)

spec.m.hv.em <- lme4::lmer(em ~ age06 + sex + educationyears +
                          time + timesq + lnwmh + hv +
                          lnwmh:timesq + hv:time + hv:lnwmh  + 
                     (1 + time|rundmcs),
                     data=df, REML=F)

spec.m.gmv.em <- lme4::lmer(em ~ age06 + sex + educationyears +
                           time + timesq + lnwmh + gmvnohv +
                           lnwmh:timesq + gmvnohv:time + gmvnohv:lnwmh +
                     (1 + time|rundmcs),
                     data=df, REML=F)

spec.m.both.em <- lme4::lmer(em ~ age06 + sex + educationyears +
                          time + timesq + lnwmh + hv + gmvnohv +
                           lnwmh:timesq + hv:time + gmvnohv:time +
                            hv:lnwmh  + gmvnohv:lnwmh +
                     (1 + time|rundmcs),
                     data=df, REML=F)

spec.m.both.em.contr <- lme4::lmer(em ~ age06 + sex + educationyears +
                          time + timesq + lnwmh + hv + gmvnohv +
                           lnwmh:timesq + hv:time + gmvnohv:time +
                     (1 + time|rundmcs),
                     data=df, REML=F)

stargazer(spec.m.wmh.wm, spec.m.hv.wm, spec.m.gmv.wm, spec.m.both.wm, spec.m.wmh.em, spec.m.hv.em, spec.m.gmv.em, spec.m.both.em,
          title = "Fixed effects results",
          ci=T, ci.level = .95, 
          omit="Constant", dep.var.labels = MemLabel,
          order=c(1,2,3,4,5,6,9,7,10,11,8,12,13),
          covariate.labels = c("Baseline age (years)", 
                               "Sex",
                               "Education (years)",
                               "Time to follow-up (linear)", 
                               "Time to follow-up (quadratic)", 
                               "WMH", 
                               "WMH progression",
                               "HV", 
                               "Hippocampal atrophy", 
                               "WMH * HV interaction",
                               "GMV",
                               "GM atrophy",
                               "WMH * GM interaction"),
          star.cutoffs = c(0.05, 0.01, 0.001),
          column.labels = c("WMH only", "WMH * HV", "WMH * GMV", "Both", "WMH only", "WMH * HV", "WMH * GMV", "Both"),
          initial.zero = F,
          header=F, type="html")

stargazer(anova(spec.m.both.wm.contr, spec.m.both.wm), summary = F, type="html")
stargazer(anova(spec.m.both.em.contr, spec.m.both.em), summary = F, type="html")

```




## 5. Determine the specificity of the effect - Other cognitive domains

We then compared this model with several alternative competing models to determine the specificity of the effect.

### 5.1 Global cognition

We wanted to test whether the effect is specific for memory, rather than general cognition.

```{r}
ci.model1 <- lme4::lmer(cognitiveindex ~ 
                       age06 + 
                       time + 
                       timesq +
                       sex + 
                       educationyears +
                     (1 + time|rundmcs),
                     data=df, REML=F)

ci.model2 <- lme4::lmer(cognitiveindex ~ age06 + time + timesq*lnwmh +
                     sex + educationyears +
                     (1 + time|rundmcs), 
                     data=df, REML=F)

ci.model3 <- lme4::lmer(cognitiveindex ~ age06 + sex + time*hv + timesq  +
                     sex + educationyears +
                     (1 + time|rundmcs),
                     data=df, REML=F)

ci.model4 <- lme4::lmer(cognitiveindex ~ age06 + time*hv + timesq*lnwmh +
                     hv*lnwmh  + sex + educationyears +
                     (1 + time|rundmcs),
                     data=df, REML=F)

ci.model5 <- lme4::lmer(cognitiveindex ~ age06 + time*hv + timesq*lnwmh +
                     sex + educationyears +
                     (1 + time|rundmcs),
                     data=df, REML=F)

stargazer(ci.model1, ci.model2, ci.model3, ci.model4, 
          title = "Fixed effects results",
          ci=T, ci.level = .95, 
          omit="Constant", dep.var.labels = "Cognitive Index",
          order=c(1,6,7,2,4,5,8,3,10,9),
          covariate.labels = c("Baseline age (years)", 
                               "Sex",
                               "Education (years)",
                               "Time to follow-up (linear)", 
                               "Time to follow-up (quadratic)", 
                               "WMH", 
                               "WMH progression",
                               "HV", 
                               "Hippocampal atrophy", 
                               "WMH * HV interaction"),
          star.cutoffs = c(0.05, 0.01, 0.001),
          column.labels = c("Null", "WMH only", "HV only", "Full"),
          initial.zero = F,
          header=F, type="html")

stargazer(anova(ci.model5, ci.model4), summary = F, type="html")


```


### 5.2 Psychomotor speed

Next, we wanted to test whether the effect is specific for memory, rather than psychomotor speed.

```{r}
ps.model1 <- lme4::lmer(psychomotorspeed ~ 
                       age06 + 
                       time + 
                       timesq +
                       sex + 
                       educationyears +
                     (1 + time|rundmcs),
                     data=df, REML=F)

ps.model2 <- lme4::lmer(psychomotorspeed ~ age06 + time + timesq*lnwmh +
                     sex + educationyears +
                     (1 + time|rundmcs), 
                     data=df, REML=F)

ps.model3 <- lme4::lmer(psychomotorspeed ~ age06 + sex + time*hv + timesq  +
                     sex + educationyears +
                     (1 + time|rundmcs),
                     data=df, REML=F)

ps.model4 <- lme4::lmer(psychomotorspeed ~ age06 + time*hv + timesq*lnwmh +
                     hv*lnwmh  + sex + educationyears +
                     (1 + time|rundmcs),
                     data=df, REML=F)

ps.model5 <- lme4::lmer(psychomotorspeed ~ age06 + time*hv + timesq*lnwmh +
                     sex + educationyears +
                     (1 + time|rundmcs),
                     data=df, REML=F)

stargazer(ps.model1, ps.model2, ps.model3, ps.model4, 
          title = "Fixed effects results",
          ci=T, ci.level = .95, 
          omit="Constant", dep.var.labels = "Psychomotor Speed",
          order=c(1,6,7,2,4,5,8,3,10,9),
          covariate.labels = c("Baseline age (years)", 
                               "Sex",
                               "Education (years)",
                               "Time to follow-up (linear)", 
                               "Time to follow-up (quadratic)", 
                               "WMH", 
                               "WMH progression",
                               "HV", 
                               "Hippocampal atrophy", 
                               "WMH * HV interaction"),
          star.cutoffs = c(0.05, 0.01, 0.001),
          column.labels = c("Null", "WMH only", "HV only", "Full"),
          initial.zero = F,
          header=F, type="html")

stargazer(anova(ps.model5, ps.model4), summary = F, type="html")

```


### 5.3 Executive function

Next, we wanted to test whether the effect is specific for memory, rather than executive function.

```{r}
ef.model1 <- lme4::lmer(executivefunction ~ 
                       age06 + 
                       time + 
                       timesq +
                       sex + 
                       educationyears +
                     (1 + time|rundmcs),
                     data=df, REML=F)

ef.model2 <- lme4::lmer(executivefunction ~ age06 + time + timesq*lnwmh +
                     sex + educationyears +
                     (1 + time|rundmcs), 
                     data=df, REML=F)

ef.model3 <- lme4::lmer(executivefunction ~ age06 + sex + time*hv + timesq  +
                     sex + educationyears +
                     (1 + time|rundmcs),
                     data=df, REML=F)

ef.model4 <- lme4::lmer(executivefunction ~ age06 + time*hv + timesq*lnwmh +
                     hv*lnwmh  + sex + educationyears +
                     (1 + time|rundmcs),
                     data=df, REML=F)

ef.model5 <- lme4::lmer(executivefunction ~ age06 + time*hv + timesq*lnwmh +
                     sex + educationyears +
                     (1 + time|rundmcs),
                     data=df, REML=F)

stargazer(ef.model1, ef.model2, ef.model3, ef.model4, 
          title = "Fixed effects results",
          ci=T, ci.level = .95, 
          omit="Constant", dep.var.labels = "Executive function",
          order=c(1,6,7,2,4,5,8,3,10,9),
          covariate.labels = c("Baseline age (years)", 
                               "Sex",
                               "Education (years)",
                               "Time to follow-up (linear)", 
                               "Time to follow-up (quadratic)", 
                               "WMH", 
                               "WMH progression",
                               "HV", 
                               "Hippocampal atrophy", 
                               "WMH * HV interaction"),
          star.cutoffs = c(0.05, 0.01, 0.001),
          column.labels = c("Null", "WMH only", "HV only", "Full"),
          initial.zero = F,
          header=F, type="html")

stargazer(anova(ef.model5, ef.model4), summary = F, type="html")

```

The combined effects of WMH and hippocampal atrophy are not specific to memory performance; they also explain global cognitive functioning as well as psychomotor speed and executive function.


[^1]: Charlton RA, Barrick TR, Markus HS, Morris RG. The relationship between episodic long-term memory and white matter integrity in normal aging. Neuropsychologia 2010; 48: 114-22.